---
title: "P8015_Hw3_agd2159"
output: github_document
author: "Zander De Jesus"
date: "10-14-2023"
---
# Problem 1: Instacart Grocery Large-Scale Data Analysis and Plotting

Begin by loading important libraries and beginning to look at full dataset. 

```{r}
library(tidyverse)
library(ggridges)
library(patchwork)
library(p8105.datasets)


data("instacart")

summary(instacart)

nrow(instacart)
ncol(instacart)

```



**Question: 1 - How many aisles are there, and which aisles are the most items ordered from?**

```{r}

num = instacart |> 
  distinct(aisle_id) |> 
  max()

aisles_df = instacart |> 
  janitor::clean_names() |> 
  group_by(aisle) |> 
  summarize(n_obs = n()) |> 
  arrange(desc(n_obs))

aisles_df

```

After creating a separate grouped tibble based on aisle name and the bulk total of orders from each aisle, we can see that the the top 3 aisles with the highest amount of items ordered are **Fresh Vegetables, Fresh Fruits, and Packaged Vegetables / Fruits.** These are then followed by Yogurt and Packaged Cheese in the dairy department. 
